Sequential rank agreement methods for comparison of ranked lists

### 2016-04-07 latex/paper.Rnw
more asymptotic

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[]{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[]{color}
\usepackage{subcaption}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{calc}
\usepackage[noae]{Sweave}
\usepackage{authblk}
\usepackage{algorithm}
\usepackage{algpseudocode}
%\usepackage{geometry}
%\geometry{a4paper, portrait, margin=1in}
%
% New commands
%
\newcommand{\ie}{\emph{i.e.}\@\xspace}
\newcommand{\eg}{\emph{e.g.}\@\xspace}

@@

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p))
\end{equation}
for a measurable non-negative real-valued distance function $f$.
For an integer $1\le d\le P$ we define the unique set of items found
by merging the first $d$ elements of each of the $L$ lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.
We then define the \emph{sequential rank agreement} as the weighted expected agreement of the items
found in the set $S_d$: [FIX: we have talked about this before, but does that redundant (L-1) factor help wrt interpretation?]
\begin{align}
\textrm{sra}(d)&= \frac{\sum_{\{p \in S_d\}}(L-1)\E[A(X_p)]}{(L-1)|S_d|}\label{def:sra}\\
   &= \sum_{\{p \in S_d\}}|S_d|^{-1} \E[A(X_p)]
\end{align}
where $|S_d|$ refers to the cardinality of the set $S_d$, and the expectation
is taken with respect to the distribution of the rankings of $X_p$ accross the $L$ lists.
Values of \textrm{sra} close to zero suggests that the lists agree while larger
values suggest disagreement. Just like in method comparison studies we
prefer smaller limits of agreement because that means that the
differences between the methods is small. If the ranked lists are
identical then the value of sequential rank agreement will be zero for
all depths $d$.

@@

Given a sample of $1,\ldots,L$ lists and estimator of estimated sequatial rank agreement is given by
\begin{equation}
\widehat{\textrm{sra}}_L(d)= \frac{\sum_{\{p \in S_d\}}(L-1)\widehat{A_L}(X_p)}{(L-1)|S_d|}\label{def:sra}
\end{equation}
where
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p)
\end{align}
is the sample variance of the rank of item $X_p$. For an easier interpretation we suggest to plot $\widehat{\textrm{sra}}_L(d)^{1/2}$ corresponding to the pooled standard deviation.

@@

\subsection{Analysis of partial/censored lists}

Incomplete or partial lists are also a common occurrence. They arise
for example in case of missing data (items), when comparing top $d$
list results from publications, or when some methods only rank a
subset of the items. For example, penalized regression based on the
Lasso provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering. Incomplete lists also occur when the analyst decides
to censor all lists at depth $d$ such that the measurement of
agreement is restricted to the first $d$ elements of each list, or
when the analyst only wants to consider the ranking of items that have
been found to be statistically significant.

@@

\section{Change-point analysis}
In many applications it is of interest to estimate a list depths which
satisfies a change point criterion. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As Hall and Schimek, we work with an infinite set of lists
$\mathcal{L}$ and study the asymptotic behaviour for
$L\to\infty$. However, we do not allow the length of the lists to
change with $L$ as in most applications the length of the lists is
fixed [FIXME: for finite $L$ are there only finitely many forskelige
lists (aber dass stoert keinen grossen geist, siger karlsson)]. First,
we define the parameter of interest. For any given threshold value $q$
a change-point $d^\ast(q)$ in the list agreement is defined by
\begin{equation}
  d^\ast(q) = \arginf_{d\in\{1,\dots,P\}}\textrm{sra}(d)>q.
\end{equation}
The value of $q$ could for example be the agreement of randomly
permuted lists [FIXME: there is only one SRA value for infinitely many
random permutations, no skygge!].
The corresponding estimated change-point is similarly defined by
\begin{equation}
\widehat{d^\ast_L}(q) = \arginf_{d\in\{1,\dots,P\}}\widehat{\textrm{sra}}_L(d)>q
\end{equation}

We start by showing that $\widehat{\textrm{sra}}_L(d)$ is consistent for $\textrm{sra}(d)$ at each
$d \in \left\{1,\ldots,P\right\}$ for $L \rightarrow \infty$. The consistency of $\widehat{d^\ast_L}(q)$
then follows by a simple corollary.

\begin{theorem}
hej
\begin{proof}
We start by noting that we can rewrite the sequential rank agreement in equation \ref{def:sra} as
\begin{equation}
\textrm{sra}(d)= \frac 1{\sum_{p=1}^P \sum_{l=1}^L 1\{R_l(X_p)\le d\}}\sum_{p=1}^P(1-\prod_{l=1}^L 1\{R_l(X_p)>d\}) \E[A(X_p)]
\end{equation}

[FIXME: Andreas details the proof with Cauchy-Schwarz.]
[FIXME: Det er lidt kedelig for saadan en fin en som Andreas: Skal theoremet og bevis ikke vaere for en general agreement metric, ikke nodvendigvis den i equation (2), og kan vi charakterisere hvad den skal opfylde?]
\end{proof}
\label{thm:consistency}
\end{theorem}

@@

The overlap among $L$ lists observed until depth $d$ is defined as the
proportion of items that are found in all $L$ lists compared to the
possible number of items found in all lists observed until depth $d$:
\begin{equation}
  O(d) = \frac{| \cap_{l=1}^L \{R_l^{-1}(r); r\leq d \} |}{d}. \label{def:overlap}
\end{equation}
Typically the overlap is only considered for pairwise comparisons
\citep{Bar-Ilan2006,Boulesteix2009} but formula \ref{def:overlap}
is easily extended to the situation where $L>2$ as shown above.

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[]{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[]{color}
\usepackage{subcaption}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{calc}
\usepackage[noae]{Sweave}
\usepackage{authblk}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
%\usepackage{geometry}
%\geometry{a4paper, portrait, margin=1in}
%
% New commands
%
\newcommand{\ie}{\emph{i.e.}\@\xspace}
\newcommand{\eg}{\emph{e.g.}\@\xspace}

@@

The agreement of the lists regarding the rank given to an item $X_p$ can be
measured by
\begin{equation}
  A(X_p) = f(R_1(X_p), \ldots, R_L(X_p))
\end{equation}
for a measurable non-negative real-valued distance function $f$.
For an integer $1\le d\le P$ we define the unique set of items found
by merging the first $d$ elements of each of the $L$ lists, i.e., the
set of items ranked less than or equal to $d$ in any of the lists:
\begin{equation}
S_d = \{R_l^{-1}(r) ; r \leq d, l = 1, \ldots, L \},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.
We then define the \emph{sequential rank agreement} as the weighted expected agreement of the items
found in the set $S_d$:
\begin{align}
\textrm{sra}(d) &= \sum_{\{p \in S_d\}}|S_d|^{-1} \E[A(X_p)]
\end{align}
where $|S_d|$ refers to the cardinality of the set $S_d$, and the expectation
is taken with respect to the distribution of the rankings of $X_p$ accross the $L$ lists.
Values of \textrm{sra} close to zero suggests that the lists agree while larger
values suggest disagreement. Just like in method comparison studies we
prefer smaller limits of agreement because that means that the
differences between the methods is small. If the ranked lists are
identical then the value of sequential rank agreement will be zero for
all depths $d$.

@@

Given a sample of $1,\ldots,L$ lists an estimator of the sequatial rank agreement is given by
\begin{equation}
\widehat{\textrm{sra}}_L(d)= \frac{\sum_{\{p \in S_d\}}(L-1)\widehat{A_L}(X_p)}{(L-1)|S_d|}\label{def:sra}
\end{equation}
where
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p)
\end{align}
is the sample variance of the rank of item $X_p$. For an easier interpretation we suggest to plot $\widehat{\textrm{sra}}_L(d)^{1/2}$ corresponding to the pooled standard deviation. The consistency of $\widehat{\textrm{sra}}_L$ as $L \rightarrow \infty$ will be discussed further in section \ref{sec:chgpoint}.

@@

\subsection{Analysis of partial/censored lists}
Incomplete or partial lists are also a common occurrence. They arise
for example in case of missing data (items), when comparing top $d$
list results from publications, or when some methods only rank a
subset of the items. For example, penalized regression based on the
Lasso provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering. Incomplete lists also occur when the analyst decides
to censor all lists at depth $d$ such that the measurement of
agreement is restricted to the first $d$ elements of each list, or
when the analyst only wants to consider the ranking of items that have
been found to be statistically significant.

@@

\section{Change-point analysis}\label{sec:chgpoint}
In many applications it is of interest to estimate a list depths which
satisfies a change point criterion. In this section we investigate the
theoretical properties of our proposed method for this specific
task. As Hall and Schimek, we work with an infinite set of lists
$\mathcal{L}$ and study the asymptotic behaviour for $L\to\infty$.
%However, we do not allow the length of the lists to
%change with $L$ as in most applications the length of the lists is
%fixed [FIXME: for finite $L$ are there only finitely many forskelige
%lists (aber dass stoert keinen grossen geist, siger karlsson)]. 
First, we define the parameter of interest. For any given threshold value $q$
a change-point $d^\ast(q)$ in the list agreement is defined by
\begin{equation}
  d^\ast(q) = \arginf_{d\in\{1,\dots,P\}}\textrm{sra}(d)>q.
\end{equation}
The value of $q$ could for example be the agreement of randomly
permuted lists [FIXME: there is only one SRA value for infinitely many
random permutations, no skygge!].
The corresponding estimated change-point is similarly defined by
\begin{equation}
\widehat{d^\ast_L}(q) = \arginf_{d\in\{1,\dots,P\}}\widehat{\textrm{sra}}_L(d)>q
\end{equation}

We start by showing that $\widehat{\textrm{sra}}_L$ is consistent for $\textrm{sra}$ for $P,L \rightarrow \infty$ and $\frac{P}{L} \rightarrow 0$. 
The consistency of $\widehat{d^\ast_L}(q)$ then follows by a simple corollary. The proof requires the following assumptions.

\begin{assumption}
We assume that:
\begin{enumerate}[label=\Alph*)]
\item{hej}
\item{med}
\item{dig}
\end{enumerate}
\end{assumption}

\begin{theorem}
Under the assumptions it follows that $\widehat{\textrm{sra}}_L \overset{P}{\longrightarrow} \textrm{sra}$ for $L,P\rightarrow \infty$ and $\frac{P}{L} \rightarrow 0$.
\begin{proof}
We start by noting that we can rewrite the sequential rank agreement in equation \ref{def:sra} as
\begin{align}
\textrm{sra}(d) &= \frac 1{\sum_{p=1}^P \sum_{l=1}^L 1\{R_l(X_p)\le d\}}\sum_{p=1}^P(1-\prod_{l=1}^L 1\{R_l(X_p)>d\}) \E[A(X_p)]\\
    &= \frac{1}{L} \sum_{l=1}^L \sum_{p=1}^P S_l(p)
\end{align}
where
\begin{align}
  S_l(p) &= ?
\end{align}
Maybe we can use the LLN for triangular arrays?
\end{proof}
\label{thm:consistency}
\end{theorem}

@@

The overlap among $L$ lists observed until depth $d$ is defined as the
proportion of items that are found in all $L$ lists compared to the
possible number of items found in all lists observed until depth $d$:
\begin{equation}
  O(d) = \frac{| \cap_{l=1}^L \{R_l^{-1}(r); r\leq d \} |}{d}\label{def:overlap}
\end{equation}
Typically the overlap is only considered for pairwise comparisons
\citep{Bar-Ilan2006,Boulesteix2009} but formula \ref{def:overlap}
is easily extended to the situation where $L>2$ as shown above.
